---
title: "Fundamentals 1"
subtitle: "Organizing numbers"
author:
  - Elizabeth King
  - Kevin Middleton
format:
  revealjs:
    theme: [default, custom.scss]
    standalone: true
    self-contained: true
    logo: QMLS_Logo.png
    slide-number: true
    show-slide-number: all
bibliography: Multivariate.bib
csl: evolution.csl
---

## Fundamentals

- Need to think about object classes more often with multivariate data and analysis
    - Use base classes: vectors, matrices, and arrays
- Every object in R has one or more classes

```{r}
#| echo: true

class(1)
class(1:5)
class(matrix(rnorm(4), nrow = 2))
class(data.frame)
```


## Scalars

- Single number (`integer` or `numeric`)
- Internally a vector of length 1

Default numeric is a *double precision floating point numeric*: $2 \times 10^{\pm308}$
```{r}
#| echo: true

x <- 4
class(x)
```

`L` specifies a *long integer*: $\pm 2 \times 10^{-9}$

```{r}
#| echo: true

x <- 4L
class(x)
```


## Vectors

- Sequence of numbers (`integer` or `numeric`)

```{r}
#| echo: true

x1 <- c(1, 2, 3, 4, 5)
class(x1)
```

```{r}
#| echo: true

x2 <- seq(1, 5, by = 1)
class(x2)
```

```{r}
#| echo: true

x3 <- 1:5
class(x3)
```

## Checking for equality

`c()` vs. `seq()`

```{r}
#| echo: true

all.equal(x1, x2)
```

`seq()` vs. `:`

```{r}
#| echo: true
all.equal(x2, x3)
identical(x2, x3)
```

Read the help for `all.equal()` for discussion about checking equality


## Matrices

- $n \times m$ object of all one data class (numeric, boolean, character)
    - *Convention*: $n$ rows by $m$ columns
- Created columnwise by default
    
```{r}
#| echo: true

matrix(1:9, nrow = 3)
matrix(1:9, nrow = 3, byrow = TRUE)
```


## Dimensions

```{r}
#| echo: true

x <- matrix(1:8, nrow = 4)
dim(x)
nrow(x)
ncol(x)
```

No `dim()`, `nrow()`, or `ncol()` for a vector

```{r}
#| echo: true

x <- 1:5
dim(x)
nrow(x)
ncol(x)
```


## Vectors are assumed to be column vectors

- Even though `ncol()` is `NULL`
- Explicitly create a row matrix if you need to

```{r}
#| echo: true

x <- 1:5
M <- matrix(x, nrow = 1)
M
dim(M)
```


## Indexing

Recall `[ ]` indexing:

```{r}
#| echo: true

M <- matrix(1:9, nrow = 3)
M[2, ]
M[, 2]
M[, 2:3]
```


## Indexing drops dimensions

Default behavior:

```{r}
#| echo: true

(M <- matrix(1:9, nrow = 3))
(x <- M[2, ])
class(x)
```


## Indexing drops dimensions

- Use `drop = FALSE` to retain dimensions
    - i.e., subset a matrix and keep it as a matrix
    - see `help("[")`: *Extract or Replace Parts of an Object*

```{r}
#| echo: true

(x <- M[2, , drop = FALSE])
class(x)
```


## Operations

Most operations work as you intuitively expect them:

```{r}
#| echo: true

x <- 1:5
x + 2
x * 2
```

Even on matrices

```{r}
#| echo: true

M <- matrix(1:4, nrow = 2)
M + 2
M * 2
```


## Element-by-element multiplication

Vectors: 

```{r}
#| echo: true

x * x
x * x * x
```

Matrices:

```{r}
#| echo: true

M * M
M * M * M
```

"Matrix multiplication" comes in the next unit


## Combining vectors and matrices

`cbind()` and `rbind()`:

```{r}
#| echo: true

rbind(x, x, x)
cbind(M, M, M)
rbind(M, M)
```


## Arrays: multidimensional matrices 

- 3 or more dimensions
- 3 dimensions is common
    - Each sample or individual has a matrix of observations (Rows, Columns)
    - Dimension 3 is samples, specimens, or individuals
- Visualize a stack of cards
    - (Rows, Columns, Samples)


## Creating arrays

```{r}
#| echo: true

M1 <- matrix(1:4, nrow = 2)
M2 <- matrix(5:8, nrow = 2)
M3 <- matrix(9:12, nrow = 2)

(AA <- array(c(M1, M2, M3), dim = c(2, 2, 3)))
```


## Naming arrays

- `dimnames()` is a list of vectors. Dimensions must match.
- Can provide on creation

```{r}
#| echo: true

dimnames(AA) <- list(c("x1", "x2"),
                     c("y1", "y2"),
                     c("M1", "M2", "M3"))
AA
```


## Naming specific dimensions

Specify the dimension using `[[]]` notation:

```{r}
#| echo: true

dimnames(AA)[[3]] <- c("m1", "m2", "m3")
AA
```


## Subsetting arrays by position

```{r}
#| echo: true

# All of element 1
AA[ , , 1]

# Column 2 of all elements
AA[ , 2 , ]

# Row 1 of all elements
AA[1 , , ]
```

May need to `t()` after


## Subsetting arrays by name

```{r}
#| echo: true

# All of m1
AA[ , , "m1"]

# Row x1 of m1
AA["x1" , , "m1"]

# Row x1 of m1 as an array
AA["x1" , , "m1", drop = FALSE]
```


## Easier arrays with `abind`

```{r}
#| echo: true

library(abind)

LL <- list(M1, M2, M3) # e.g., after using map() to read a set of files
abind(LL, along = 3)
```

