---
title: "MANOVA"
author:
  - Elizabeth King
  - Kevin Middleton
format:
  revealjs:
    theme: [default, custom.scss]
    standalone: true
    self-contained: true
    logo: QMLS_Logo.png
    slide-number: true
    show-slide-number: all
bibliography: Multivariate.bib
csl: evolution.csl
---

## Multivariate ANOVA

```{r}
#| label: setup
#| echo: false
#| warning: false
#| message: false

library(tidyverse)
library(cowplot)
library(plotly)

ggplot2::theme_set(theme_cowplot())
```

How well can a categorical predictor variable jointly predict a multivariate set of outcome variables?

**MANOVA**:

`Height + Mouth_Diam + Tube_Diam + Keel_Diam ~ Site`

## MANOVA creates a composite variable

- In PCA, a composite is created that explains the maximum amount of variance among the variables
- In MANOVA, a composite is created that does this while also maximally separating the groups in the categorical predictor
    - Then this composite is used as the dependent variable in an ANOVA
- Focus is on mean differences and whether the combined differences are likely to have occurred by chance

## Advantages of MANOVA

- Identify associations not clear from individual ANOVAs
- False positive correction
- More variables often means more information

## Assumptions and Limitations of MANOVA

- Same as ANOVA but for everything
- Multivariate normal
- Equal variances
- Relationships between variables

## Data

```{r}

nn <- 25
dat <- tibble(y1 = c(rnorm(nn),rnorm(nn,1)),
              y2 = c(rnorm(nn),rnorm(nn,1)),
              y3 = c(rnorm(nn),rnorm(nn,1)),
              xx = rep(c("A","B"), each = nn))

fig <- plot_ly() |>
  add_markers(data = dat,
              x = ~ y1,
              y = ~ y2,
              z = ~ y3,
              color = ~ xx,
              marker = list(size = 5),
              showlegend = FALSE) |>
  layout(scene = list(xaxis = list(title = 'y1'),
                      yaxis = list(title = 'y2'),
                      zaxis = list(title = 'y3')))

fig
```


## References

::: {#refs}
:::
