---
title: "Canonical Ordination 2"
subtitle: "Symmetrical"
author:
  - Elizabeth King
  - Kevin Middleton
format:
  revealjs:
    theme: [default, custom.scss]
    standalone: true
    self-contained: true
    logo: QMLS_Logo.png
    slide-number: true
    show-slide-number: all
bibliography: Multivariate.bib
csl: evolution.csl
---


```{r}
#| label: setup
#| echo: false
#| warning: false
#| message: false

library(tidyverse)
library(cowplot)

ggplot2::theme_set(theme_cowplot())

```

## Symmetrical ordinations

- All matrices have the same role
- No outcome vs. predictor relationship
- No hypotheses of causation
- $M_1 \sim M_2 == M_2 \sim M_1$


## Symmetrical ordinations

1. Canonical Correlation Analysis (CCorA)
2. Co-inertia Analysis (CoIA)
3. Multiple factor analysis (MFA)


## Canonical Correlation Analysis

- Two sets of data
- Align canonical axes to maximize correlations
- Data requirements
    - Multivariate normality
    - No categorical variables


## Doubs River environmental data

```{r}
load("Data/Doubs.RData")

spe <- spe[-8, ]
env <- env[-8, ]
spa <- spa[-8, ]

# Split off dfs. env2 doesnt have dfs 
env2 <- env[, -1]
```

```{r}
#| echo: true

Topo <- env2[, c(1:3)]
names(Topo)

Chem <- env2[, c(4:10)]
names(Chem)

```


## Density plots

```{r}
#| echo: true
#| output-location: slide

Topo |> 
  pivot_longer(cols = everything()) |> 
  ggplot(aes(value)) +
  geom_density() +
  facet_grid(name ~ ., scales = "free")
```


## References

::: {#refs}
:::

